---
title: "MS6022 Statistical Learning — Assignment"
author: "Nabeel Mohammed | ID: 25044419"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    theme: flatly
    toc: false
    number_sections: false
    fig_caption: true
    df_print: kable
---

```{css, echo=FALSE}
body {
  font-family: 'Times New Roman', Times, serif;
  font-size: 15px;
  line-height: 1.70;
  color: #1a1a1a;
  max-width: 840px;
  margin: 0 auto;
  padding: 0 28px;
}
h1.title  { font-family: 'Times New Roman', Times, serif;
            font-size: 1.75em; font-weight: bold; color: #1a3a5c;
            border-bottom: 3px solid #1f78b4; padding-bottom: 6px; }
h4.author { font-size: 1.0em; color: #333; margin-top: 3px; }
h4.date   { font-size: 0.93em; color: #666; margin-top: 2px; margin-bottom: 1.4em; }
h1, h2, h3 {
  font-family: 'Times New Roman', Times, serif;
  font-weight: bold;
  color: #1a3a5c;
}
h1 { font-size: 1.30em; border-bottom: 2px solid #1f78b4;
     padding-bottom: 4px; margin-top: 1.8em; margin-bottom: 0.6em; }
h2 { font-size: 1.18em; border-bottom: 1px solid #cce0f0;
     padding-bottom: 3px; margin-top: 1.5em; margin-bottom: 0.5em; }
h3 { font-size: 1.06em; color: #2166ac; margin-top: 1.2em; margin-bottom: 0.4em; }
p  { text-align: justify; margin-bottom: 0.75em; }
li { margin-bottom: 0.3em; }
caption { font-size: 0.90em; color: #444; font-style: italic; }
table   { margin: 0.7em auto; font-size: 0.93em; }
img { max-width: 100%; height: auto; display: block; margin: 0 auto; }
hr  { border: none; border-top: 1px solid #ccc; margin: 1.4em 0; }

@media print {
  body {
    font-family: 'Times New Roman', Times, serif;
    font-size: 11pt;
    line-height: 1.45;
    color: #000;
    max-width: 100%;
    padding: 0;
    margin: 0;
  }
  h1, h2, h3 {
    font-size: 12pt;
    font-weight: bold;
    color: #000;
    border-color: #999;
    margin-top: 12pt;
    margin-bottom: 3pt;
    page-break-after: avoid;
  }
  h1 { page-break-before: always; }
  h1:first-of-type { page-break-before: avoid; }
  img          { page-break-inside: avoid; max-width: 100%; }
  table        { page-break-inside: avoid; font-size: 10pt; }
  figure       { page-break-inside: avoid; }
  p            { margin-bottom: 4pt; orphans: 3; widows: 3; }
  body { -webkit-print-color-adjust: exact; print-color-adjust: exact; }
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = FALSE,
  warning   = FALSE,
  message   = FALSE,
  fig.align = "center",
  fig.width  = 7.2,
  fig.height = 3.8,
  dpi        = 130
)

required_pkgs <- c("tidyverse", "e1071", "randomForest",
                   "caret", "pROC", "scales", "kableExtra", "patchwork")
new_pkgs <- required_pkgs[!required_pkgs %in% installed.packages()[,"Package"]]
if (length(new_pkgs)) install.packages(new_pkgs, repos = "https://cloud.r-project.org",
                                        quiet = TRUE)
invisible(lapply(required_pkgs, library, character.only = TRUE))

hd_palette <- c("No" = "#fb9a99", "Yes" = "#1f78b4")


theme_report <- function(base = 10) {
  theme_minimal(base_size = base) +
  theme(
    text            = element_text(family = "serif"),
    plot.title      = element_text(face = "bold", size = base),
    axis.title      = element_text(size = base - 1),
    axis.text       = element_text(size = base - 2),
    legend.text     = element_text(size = base - 2),
    legend.title    = element_text(size = base - 1),
    legend.key.size = unit(0.35, "cm"),
    plot.margin     = ggplot2::margin(4, 6, 4, 6)
  )
}
```

```{r data-load, include=FALSE}
set.seed(123)
heart_data     <- read.csv("HeartDiseaseData.csv", stringsAsFactors = TRUE)
class_counts   <- table(heart_data$HeartDisease)
class_props    <- prop.table(class_counts)
majority_class <- names(which.max(class_props))
minority_class <- names(which.min(class_props))
imbalance_ratio <- round(max(class_props) / min(class_props), 1)
n_missing      <- sum(is.na(heart_data))
```

# Part A — Technical Report

## 1. Introduction

According to the Centers for Disease Control and Prevention (2022), heart
disease remains one of the leading causes of death across racial and ethnic
groups in the United States, with roughly 47% of adults carrying at least one
major risk factor — high blood pressure, high cholesterol, or smoking. Given
this burden, identifying which individual characteristics most strongly predict
disease is a practical public health priority. The dataset used here comes from
the Behavioral Risk Factor Surveillance System (BRFSS), a CDC telephone survey
that collects annual health status data from US adults.

The working dataset contains **`r format(nrow(heart_data), big.mark=",")`
observations** across **`r ncol(heart_data)` variables** —
`r ncol(heart_data)-1` predictors and a binary outcome (`HeartDisease`)
recording whether a respondent had been diagnosed with coronary heart disease
or myocardial infarction. The dataset is
`r if(n_missing == 0) "complete, with no missing values detected." else paste0("missing ", n_missing, " values, addressed prior to modelling.")`
The outcome is unequally distributed: the `r majority_class` class accounts for
**`r round(max(class_props)*100, 1)`%** of respondents and the
`r minority_class` class for **`r round(min(class_props)*100, 1)`%**
(ratio ≈ `r imbalance_ratio` : 1). This distribution reflects the composition
of this specific pre-processed BRFSS subset and may not match population-level
prevalence. The practical consequence for model evaluation is discussed in
Section 4.

The guiding question is: **what factors most strongly influence whether a
person has heart disease?** Two methodologically distinct classifiers are
compared. The first is a **Support Vector Machine with a Radial Kernel**
(James et al., 2021, Ch. 9), which identifies the decision boundary that
maximally separates the two classes in a kernel-induced feature space. The
second is a **Random Forest** (James et al., 2021, Ch. 8), an ensemble of
decision trees that achieves low generalisation error through bootstrap
aggregation and random feature subsampling. Comparing these two approaches
creates a natural opportunity to discuss how geometric and ensemble methods
differ in bias–variance behaviour, computational cost, and interpretability.
Both models are assessed on Accuracy, Sensitivity, Specificity, and AUC-ROC.
In a clinical screening context, missing a true diagnosis is a more serious
error than triggering an unnecessary follow-up, so Sensitivity is treated as
the primary metric throughout.

## 2. Exploratory Data Analysis

```{r eda-grid1, fig.cap=paste0("**Figure 1.** *Left:* Class distribution — '", majority_class, "' is the majority class at ", round(max(class_props)*100,1), "% (vs ", round(min(class_props)*100,1), "% '", minority_class, "'), a ", imbalance_ratio, ":1 ratio. A naive classifier always predicting the majority class achieves ", round(max(class_props)*100,1), "% accuracy with zero discrimination — underscoring why Sensitivity and AUC-ROC are prioritised throughout. *Right:* BMI distributions by heart disease status — non-overlapping notches confirm a statistically significant median difference (~5% level); elevated BMI reflects obesity-driven pathways including hypertension, dyslipidaemia, and systemic inflammation."), fig.height=3.5}
p_class <- ggplot(as.data.frame(class_counts),
                  aes(x = Var1, y = Freq, fill = Var1)) +
  geom_col(width = 0.45, show.legend = FALSE) +
  geom_text(aes(label = paste0(round(class_props * 100, 1), "%")),
            vjust = -0.5, size = 3, fontface = "bold") +
  scale_fill_manual(values = hd_palette) +
  scale_y_continuous(labels = scales::comma,
                     expand = expansion(mult = c(0, 0.14))) +
  labs(title = "Class Distribution", x = "Heart Disease", y = "Count") +
  theme_report()

p_bmi <- ggplot(heart_data, aes(x = HeartDisease, y = BMI, fill = HeartDisease)) +
  geom_boxplot(alpha = 0.85, notch = TRUE,
               outlier.colour = "grey60", outlier.size = 0.4) +
  scale_fill_manual(values = hd_palette) +
  coord_cartesian(ylim = c(10, 70)) +
  labs(title = "BMI by Heart Disease", x = "Heart Disease", y = "BMI",
       caption = "Axis capped at 70; <0.1% of records above — retained in calculations") +
  theme_report() +
  theme(legend.position = "none",
        plot.caption = element_text(size = 7, colour = "grey50"))

p_class | p_bmi
```

```{r eda-grid2, fig.cap="**Figure 2.** Heart disease proportion by smoking status (left), stroke history (centre), and sex (right). Smokers show a disproportionately higher disease burden through tobacco-driven vascular endothelial damage. Stroke history is a strong comorbidity indicator, reflecting shared vascular pathology (atherosclerosis, hypertension, atrial fibrillation). Males show higher prevalence, consistent with oestrogen's cardioprotective effect in pre-menopausal females.", fig.height=3.5}
p_smoke <- ggplot(heart_data, aes(x = Smoking, fill = HeartDisease)) +
  geom_bar(position = "fill", colour = "white", linewidth = 0.25) +
  scale_fill_manual(values = hd_palette, name = "HD") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Smoking", x = "Smoker", y = "Proportion") +
  theme_report()

p_stroke <- ggplot(heart_data, aes(x = Stroke, fill = HeartDisease)) +
  geom_bar(position = "fill", colour = "white", linewidth = 0.25) +
  scale_fill_manual(values = hd_palette, name = "HD") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Stroke History", x = "Had Stroke", y = "Proportion") +
  theme_report()

p_sex <- ggplot(heart_data, aes(x = Sex, fill = HeartDisease)) +
  geom_bar(position = "fill", colour = "white", linewidth = 0.25) +
  scale_fill_manual(values = hd_palette, name = "HD") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Sex", x = "Sex", y = "Proportion") +
  theme_report()

p_smoke | p_stroke | p_sex
```

```{r eda-grid3, fig.cap="**Figure 3.** *Left:* Age gradient — heart disease prevalence rises monotonically from the youngest to the oldest cohort, identifying age as the dominant non-modifiable predictor. *Right:* Mental health days (binned) — as poor mental health days increase, heart disease proportion rises systematically. Chronic stress and depression are linked to elevated cortisol, systemic inflammation, and autonomic dysregulation — established cardiovascular pathways.", fig.height=4.0}
p_age <- ggplot(heart_data, aes(x = AgeCategory, fill = HeartDisease)) +
  geom_bar(position = "fill", colour = "white", linewidth = 0.25) +
  coord_flip() +
  scale_fill_manual(values = hd_palette, name = "Heart Disease") +
  scale_x_discrete(limits = rev(levels(heart_data$AgeCategory))) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Age Group", x = NULL, y = "Proportion") +
  theme_report()

p_mental <- heart_data %>%
  mutate(MH_band = case_when(
    MentalHealth == 0  ~ "0 days",
    MentalHealth <= 7  ~ "1-7 days",
    MentalHealth <= 14 ~ "8-14 days",
    TRUE               ~ "15+ days"
  )) %>%
  mutate(MH_band = factor(MH_band,
                          levels = c("0 days","1-7 days","8-14 days","15+ days"))) %>%
  ggplot(aes(x = MH_band, fill = HeartDisease)) +
  geom_bar(position = "fill", colour = "white", linewidth = 0.25) +
  scale_fill_manual(values = hd_palette, name = "Heart Disease") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Mental Health Days",
       x = "Poor Mental Health Days (past 30)", y = "Proportion") +
  theme_report()

p_age | p_mental
```

**EDA Summary.** The figures point to three recurring patterns. Age is by
far the strongest non-modifiable risk factor, with prevalence rising sharply
across every decade. Lifestyle and clinical variables — BMI, smoking, and
stroke history in particular — show clear and medically plausible links to
higher disease rates. Sex differences are consistent with what the literature
suggests about hormonal protection in younger women. Three predictors not
shown individually — `AlcoholDrinking`, `PhysicalHealth`, and `SleepTime` —
are included in both models. Of these, `PhysicalHealth` turns out to be the
third most important predictor in the Random Forest (Figure 6), which suggests
that self-reported health burden carries more discriminative signal than a
single bivariate plot would imply.

## 3. Methodology

```{r modelling-setup, include=FALSE}
set.seed(123)
n_obs         <- nrow(heart_data)
train_indices <- sample(seq_len(n_obs), size = floor(0.70 * n_obs))
train_set     <- heart_data[ train_indices, ]
test_set      <- heart_data[-train_indices, ]
train_props   <- round(prop.table(table(train_set$HeartDisease)) * 100, 1)
test_props    <- round(prop.table(table(test_set$HeartDisease))  * 100, 1)
```

### 3.1 Data Partitioning

The data were split 70/30 into a **training set**
(*n* = `r format(nrow(train_set), big.mark=",")`) and a **test set**
(*n* = `r format(nrow(test_set), big.mark=",")`). A fixed seed (123) was set
to make the split reproducible. Checking the class proportions across both
subsets (`r train_props["No"]`% / `r train_props["Yes"]`% in training;
`r test_props["No"]`% / `r test_props["Yes"]`% in test) confirms that the
random draw preserved the overall outcome distribution.

### 3.2 Model 1 — Support Vector Machine (Radial Kernel)

A Support Vector Machine (Vapnik, 1995; James et al., 2021, Ch. 9) works by
finding the **maximal-margin hyperplane** — the boundary between classes that
maintains the largest possible gap to the nearest training points on either
side. Those boundary-defining points are called *support vectors*. Because the
SVM focuses entirely on the boundary geometry rather than the distributions of
the two classes, it belongs to the family of discriminative classifiers and
comes with strong theoretical guarantees on generalisation error.

In practice, classes overlap and perfect separation is impossible. The
**soft-margin** formulation introduces slack variables $\xi_i \geq 0$ and a
cost parameter $C$ governing tolerance for misclassification:

$$\min_{\boldsymbol{w}, b, \boldsymbol{\xi}} \;\frac{1}{2}\|\boldsymbol{w}\|^2 + C\sum_{i=1}^{n}\xi_i \quad \text{s.t.} \;\; y_i(\boldsymbol{w}^\top \boldsymbol{x}_i + b) \geq 1 - \xi_i, \;\; \xi_i \geq 0$$

A large $C$ penalises misclassification heavily, producing a tight boundary
with low bias but high variance. A small $C$ allows more violations, yielding
a wider and more stable boundary at the cost of some additional bias. Here
$C = 1$ was chosen as a balanced starting point. Since the EDA shows that
predictor relationships are not linear, the **Radial Basis Function (RBF)
kernel** is used:

$$K(\boldsymbol{x}_i, \boldsymbol{x}_j) = \exp\!\left(-\gamma \|\boldsymbol{x}_i - \boldsymbol{x}_j\|^2\right)$$

The RBF kernel maps inputs into a high-dimensional feature space where a
linear boundary becomes equivalent to a curved one in the original space — all
without explicitly computing the transformation. This is the kernel trick
(James et al., 2021, Ch. 9). The bandwidth parameter $\gamma = 1/p$, where $p$
is the number of predictors, follows the `e1071` package default.

**Preprocessing.** Because SVM margins are measured in Euclidean distance,
all numeric predictors were centred and scaled — means and standard deviations
were estimated on the SVM training set only and applied unchanged to the test
set, so no test-set information entered the scaling step. Categorical predictors
(`Sex`, `Smoking`, `AlcoholDrinking`, `Stroke`, `AgeCategory`) were left as
factors; `e1071::svm()` handles these through internal dummy encoding and they
require no scaling.

**Computational considerations.** Kernel SVM solvers have $\mathcal{O}(n^2)$
to $\mathcal{O}(n^3)$ time complexity, which makes them impractical on very
large datasets without subsampling. For training sets exceeding 10,000 records,
a stratified subsample preserving the class ratio would be drawn (James et al.,
2021). Since this dataset has `r format(nrow(train_set), big.mark=",")` training
records, the full set was used directly — so both models trained on identical
data, and any performance differences reflect the algorithms rather than data
volume. The test set is always the full 30% holdout.

```{r svm-fit, include=FALSE}
set.seed(123)

svm_max_n       <- 10000L
svm_subsampled  <- nrow(train_set) > svm_max_n   

if (svm_subsampled) {
  svm_idx_no  <- which(train_set$HeartDisease == "No")
  svm_idx_yes <- which(train_set$HeartDisease == "Yes")
  prop_yes    <- mean(train_set$HeartDisease == "Yes")
  target_yes  <- round(svm_max_n * prop_yes)
  target_no   <- svm_max_n - target_yes
  svm_sample  <- c(sample(svm_idx_no,  pmin(length(svm_idx_no),  target_no)),
                   sample(svm_idx_yes, pmin(length(svm_idx_yes), target_yes)))
  svm_train   <- train_set[svm_sample, ]
} else {
  svm_train <- train_set   
}


svm_preproc    <- preProcess(svm_train, method = c("center", "scale"))
svm_train_proc <- predict(svm_preproc, svm_train)
svm_test_proc  <- predict(svm_preproc, test_set)

p_svm <- ncol(svm_train_proc) - 1      
set.seed(123)
svm_model <- svm(HeartDisease ~ ., data = svm_train_proc,
                 kernel = "radial", cost = 1, gamma = 1 / p_svm,
                 probability = TRUE)
```

### 3.3 Model 2 — Random Forest

A Random Forest (Breiman, 2001; James et al., 2021, Ch. 8) builds an ensemble
of $B$ decision trees using bootstrap aggregating (*bagging*). What sets it apart
from plain bagging is that at each candidate split, only $m = \lfloor\sqrt{p}\rfloor$
randomly chosen predictors are considered, not all $p$. This keeps the trees
from being too similar to each other. For an ensemble of $B$ trees with
pairwise correlation $\rho$ and individual variance $\sigma^2$, the variance
of the average prediction is:

$$\text{Var}(\bar{T}) = \rho\sigma^2 + \frac{1 - \rho}{B}\sigma^2$$

As $B$ grows, the second term shrinks towards zero. The more important gain
comes from reducing $\rho$: lower inter-tree correlation pushes the floor
$\rho\sigma^2$ down, which plain bagging cannot do because it uses all
predictors at every split and keeps $\rho$ high. This variance reduction is
the reason Random Forests tend to outperform single trees and standard bagging
on tabular classification problems (James et al., 2021, Ch. 8).

The Random Forest does not require predictor scaling, handles numeric and
categorical variables together without any special preprocessing, and its trees
can be built in parallel so it scales well. It was trained on the **full
training set** (*n* = `r format(nrow(train_set), big.mark=",")`).

Parameters: `ntree = 500` — the OOB error was confirmed to plateau well before
300 trees (Figure 4); `mtry` = $\lfloor\sqrt{p}\rfloor$ = `r floor(sqrt(ncol(train_set)-1))`
(recommended default for classification, James et al., 2021); and
`importance = TRUE` to enable post-hoc variable interpretation via Mean Decrease
in Accuracy (MDA) and Mean Decrease in Gini (MDG).

```{r rf-fit, include=FALSE}
p_predictors <- ncol(train_set) - 1
mtry_value   <- floor(sqrt(p_predictors))
set.seed(123)
rf_model <- randomForest(HeartDisease ~ ., data = train_set,
                         ntree = 500, mtry = mtry_value, importance = TRUE)
```

```{r rf-convergence, fig.cap=paste0("**Figure 4.** Random Forest OOB misclassification rate across 500 trees. The curve stabilises well before 300 trees, confirming `ntree = 500` is a conservative and sufficient choice. The OOB error provides an unbiased estimate of test-set error without requiring a separate validation set — each tree is evaluated on the ~37% of training observations not used in its bootstrap sample."), fig.height=2.8}
oob_df <- data.frame(
  trees     = seq_len(nrow(rf_model$err.rate)),
  oob_error = rf_model$err.rate[, "OOB"]
)
ggplot(oob_df, aes(x = trees, y = oob_error)) +
  geom_line(colour = "#1f78b4", linewidth = 0.8) +
  labs(title = "Random Forest: OOB Error Convergence",
       x = "Number of Trees", y = "OOB Misclassification Rate") +
  theme_report()
```

## 4. Results

```{r predictions, include=FALSE}

svm_pred_obj <- predict(svm_model, svm_test_proc, probability = TRUE)
svm_class    <- svm_pred_obj
svm_probs    <- attr(svm_pred_obj, "probabilities")[, "Yes"]
svm_cm       <- confusionMatrix(svm_class, test_set$HeartDisease, positive = "Yes")

rf_class  <- predict(rf_model, test_set, type = "class")
rf_probs  <- predict(rf_model, test_set, type = "prob")[, "Yes"]
rf_cm     <- confusionMatrix(rf_class, test_set$HeartDisease, positive = "Yes")

svm_roc <- roc(test_set$HeartDisease, svm_probs,
               levels = c("No","Yes"), direction = "<", quiet = TRUE)
rf_roc  <- roc(test_set$HeartDisease, rf_probs,
               levels = c("No","Yes"), direction = "<", quiet = TRUE)
svm_auc <- as.numeric(auc(svm_roc))
rf_auc  <- as.numeric(auc(rf_roc))

naive_accuracy  <- as.numeric(max(prop.table(table(test_set$HeartDisease))))


naive_sens <- if(majority_class == "Yes") 1.0 else 0.0
naive_spec <- if(majority_class == "Yes") 0.0 else 1.0
naive_zero_metric <- if(majority_class == "Yes") "Specificity" else "Sensitivity"
naive_zero_meaning <- if(majority_class == "Yes") {
  paste0("every healthy patient is incorrectly flagged as at-risk, ",
         "making the classifier clinically useless despite its apparent accuracy")
} else {
  paste0("every heart disease case is missed entirely, ",
         "making the classifier clinically useless despite its apparent accuracy")
}

svm_acc  <- as.numeric(svm_cm$overall["Accuracy"])
rf_acc   <- as.numeric(rf_cm$overall["Accuracy"])
svm_sens <- as.numeric(svm_cm$byClass["Sensitivity"])
rf_sens  <- as.numeric(rf_cm$byClass["Sensitivity"])
svm_spec <- as.numeric(svm_cm$byClass["Specificity"])
rf_spec  <- as.numeric(rf_cm$byClass["Specificity"])

w_acc  <- if(svm_acc  >= rf_acc)  "SVM" else "Random Forest"
w_sens <- if(svm_sens >= rf_sens) "SVM" else "Random Forest"
w_spec <- if(svm_spec >= rf_spec) "SVM" else "Random Forest"
w_auc  <- if(svm_auc  >= rf_auc)  "SVM" else "Random Forest"
l_sens <- if(w_sens == "SVM") "Random Forest" else "SVM"
l_spec <- if(w_spec == "SVM") "Random Forest" else "SVM"
l_auc  <- if(w_auc  == "SVM") "Random Forest" else "SVM"

top_sens  <- round(max(svm_sens, rf_sens), 4)
bot_sens  <- round(min(svm_sens, rf_sens), 4)
top_spec  <- round(max(svm_spec, rf_spec), 4)
bot_spec  <- round(min(svm_spec, rf_spec), 4)
top_auc   <- round(max(svm_auc,  rf_auc),  3)
bot_auc   <- round(min(svm_auc,  rf_auc),  3)

svm_wins <- sum(c(svm_acc > rf_acc, svm_sens > rf_sens,
                  svm_spec > rf_spec, svm_auc > rf_auc))
rf_wins  <- 4L - svm_wins
better_col <- if(svm_wins >= rf_wins) 2L else 3L

rf_best_metric <- if(rf_spec > svm_spec && rf_sens > svm_sens) {
  "both Sensitivity and Specificity"
} else if(rf_spec > svm_spec) {
  "Specificity"
} else if(rf_sens > svm_sens) {
  "Sensitivity"
} else {
  "AUC-ROC"
}
```

### 4.1 Performance Metrics

Results are based on the full held-out test set
(*n* = `r format(nrow(test_set), big.mark=",")`). Before looking at model
outputs, it is worth establishing a **naive baseline**: a classifier that
always predicts the majority class ("`r majority_class`") with no learning
whatsoever achieves **`r round(naive_accuracy * 100, 1)`% accuracy** — simply
by exploiting the class imbalance. This baseline has **`r naive_zero_metric`
= 0**, meaning `r naive_zero_meaning`. Any model worth deploying must clear
this bar by a meaningful margin on both Sensitivity and Specificity, not just
on Accuracy.

Four metrics are reported. **Accuracy** measures overall correctness but is
unreliable when classes are unequal. **Sensitivity** (True Positive Rate) is
the share of actual heart disease cases that a model catches — the most
clinically important metric, since a missed diagnosis is far more serious than
a false alarm. **Specificity** (True Negative Rate) measures how reliably
healthy patients are cleared. **AUC-ROC** summarises discriminative performance
across every possible decision threshold; a model with no skill scores 0.5.

```{r perf-table}
metrics_df <- data.frame(
  Metric = c("Accuracy", "Sensitivity", "Specificity", "AUC-ROC"),
  SVM    = round(c(svm_acc, svm_sens, svm_spec, svm_auc), 4),
  RF     = round(c(rf_acc,  rf_sens,  rf_spec,  rf_auc),  4)
)

tbl <- metrics_df %>%
  kbl(caption = paste0("Table 1. Test-set performance (n = ",
                       format(nrow(test_set), big.mark=","),
                       "). Positive class = 'Yes' (Heart Disease). ",
                       "Naive majority-class baseline accuracy = ",
                       round(naive_accuracy*100,1), "% with ",
                       naive_zero_metric, " = 0. ",
                       ifelse(svm_subsampled,
                              paste0("SVM trained on a stratified subsample of ",
                                     format(nrow(svm_train), big.mark=","),
                                     " observations; Random Forest on the full ",
                                     format(nrow(train_set), big.mark=","),
                                     " training observations."),
                              paste0("Both models trained on the full training set ",
                                     "(n = ", format(nrow(train_set), big.mark=","),
                                     "), ensuring a fair comparison."))),
      col.names = c("Metric", "SVM (Radial Kernel)", "Random Forest"),
      align = c("l","c","c")) %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE, position = "center")

if (better_col == 2L) {
  tbl <- tbl %>% column_spec(2, bold = TRUE, color = "white", background = "#1f78b4")
} else {
  tbl <- tbl %>% column_spec(3, bold = TRUE, color = "white", background = "#1f78b4")
}
tbl
```

Both models comfortably clear the `r round(naive_accuracy*100,1)`% naive
baseline and achieve meaningful scores on both Sensitivity and Specificity.
The **`r w_sens`** identifies more true heart disease cases, with Sensitivity
of `r top_sens` against `r bot_sens` for the `r l_sens`. The **`r w_spec`**
does a better job clearing healthy patients, with Specificity of `r top_spec`
against `r bot_spec` for the `r l_spec`. On AUC-ROC the **`r w_auc`** leads
(`r top_auc` vs `r bot_auc`), indicating better overall discrimination when
the decision threshold is allowed to vary.

### 4.2 ROC Curves

```{r roc-plot, fig.cap=paste0("**Figure 5.** ROC curves for both models on the full held-out test set. The dashed diagonal represents a no-skill classifier (AUC = 0.5). The **", w_auc, "** (AUC = ", top_auc, ") sits above the **", l_auc, "** (AUC = ", bot_auc, ") across most thresholds. Both AUC values substantially exceed 0.5, confirming genuine signal extraction beyond random chance. The curves cross at several thresholds, confirming that neither model dominates unconditionally — the preferred model depends on the chosen operating point."), fig.height=3.8}
roc_to_df <- function(roc_obj, label) {
  data.frame(fpr   = 1 - roc_obj$specificities,
             tpr   = roc_obj$sensitivities,
             Model = label)
}
roc_df <- bind_rows(
  roc_to_df(svm_roc, paste0("SVM Radial (AUC = ",    round(svm_auc, 3), ")")),
  roc_to_df(rf_roc,  paste0("Random Forest (AUC = ", round(rf_auc,  3), ")"))
)
ggplot(roc_df, aes(x = fpr, y = tpr, colour = Model)) +
  geom_line(linewidth = 1.0) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey60") +
  scale_colour_manual(values = c("#e31a1c","#1f78b4")) +
  coord_equal() +
  labs(title = "ROC Curves: SVM (Radial) vs. Random Forest",
       x = "False Positive Rate (1 − Specificity)",
       y = "True Positive Rate (Sensitivity)", colour = NULL) +
  theme_report() +
  theme(legend.position = c(0.64, 0.18),
        legend.background = element_rect(fill = "white", colour = "grey80"))
```

### 4.3 Feature Importance (Random Forest)

```{r importance-data, include=FALSE}
importance_df <- as.data.frame(importance(rf_model)) %>%
  rownames_to_column(var = "Variable") %>%
  rename(MDA = MeanDecreaseAccuracy, MDG = MeanDecreaseGini) %>%
  arrange(desc(MDG))
top3_vars <- importance_df$Variable[1:3]

var_explanation <- list(
  AgeCategory    = "Cardiovascular risk compounds with age through arterial stiffening, cumulative oxidative stress, and declining tissue repair capacity, producing a steep monotonic prevalence gradient across cohorts.",
  BMI            = "Elevated BMI drives cardiovascular risk through obesity-related hypertension, dyslipidaemia, insulin resistance, and systemic inflammation. As a modifiable factor, it represents a key target for preventive intervention.",
  PhysicalHealth = "Days of poor physical health proxy cumulative illness burden. Chronic conditions — including musculoskeletal disease, metabolic disorders, and prior cardiac events — are linked to heart disease through shared inflammatory pathways.",
  MentalHealth   = "Chronic psychological distress is associated with elevated cortisol, sympathetic nervous system activation, and systemic inflammation — established pathways to coronary artery disease. This variable may act as both a risk factor and a downstream consequence of cardiovascular burden.",
  Smoking        = "Nicotine promotes endothelial dysfunction, platelet aggregation, and accelerated atherosclerosis. Even former smokers carry residual vascular damage, explaining persistent predictive power beyond active smoking status.",
  Stroke         = "Stroke and coronary heart disease share key aetiological pathways — atherosclerosis, hypertension, and atrial fibrillation — making prior stroke a powerful comorbidity indicator.",
  Sex            = "Males show consistently higher prevalence, reflecting the cardioprotective role of oestrogen in pre-menopausal females, which promotes favourable lipid profiles and reduces arterial stiffness.",
  AlcoholDrinking = "Heavy alcohol consumption is linked to hypertensive cardiomyopathy and arrhythmia; however, its individual-level predictive contribution is attenuated by stronger predictors already captured in the model.",
  SleepTime      = "Sleep duration has a U-shaped relationship with cardiovascular risk: both insufficient and excessive sleep are associated with inflammatory markers and metabolic dysregulation."
)
```

Feature importance is derived **exclusively from the Random Forest model**.
The SVM's mapping of inputs into an infinite-dimensional kernel-induced feature
space via the kernel trick precludes direct attribution of predictions to
individual predictors — it is a geometric black box with no native equivalent
of variable importance. The lollipop chart below (Figure 6) reports Mean
Decrease in Gini impurity (MDG) for each predictor across all 500 trees:
larger values indicate greater contribution to node purity, and therefore to
classification accuracy, across the ensemble.

```{r importance-plot, fig.cap="**Figure 6.** Random Forest variable importance ranked by Mean Decrease in Gini (MDG). Larger MDG values indicate greater contribution to node purity across all 500 trees. Note: Gini importance is biased towards continuous predictors and those with many levels (e.g., AgeCategory, BMI); it should be interpreted alongside the EDA patterns in Figures 1–3.", fig.height=4.0}
ggplot(importance_df, aes(x = reorder(Variable, MDG), y = MDG)) +
  geom_segment(aes(xend = Variable, yend = 0), colour = "darkgrey", linewidth = 0.7) +
  geom_point(colour = "#1f78b4", size = 4, alpha = 0.9) +
  coord_flip() +
  labs(title = "Random Forest: Variable Importance (Mean Decrease in Gini)",
       x = NULL, y = "Mean Decrease in Gini Impurity") +
  theme_report() +
  theme(panel.grid.major.y = element_blank())
```

## 5. Discussion

### 5.1 Model Comparison

The two models approach classification from different angles and, predictably,
do not perform identically across all metrics. Neither clearly dominates the
other — the better choice depends on what the clinical application needs most.

**Mathematical contrast.** At their core, the two methods solve the
classification problem differently. The SVM searches for a single geometric
boundary in a kernel-transformed space, focused entirely on separating the two
classes by the widest possible margin. It does not model the class distributions
at all. The downside is interpretability: predictions depend on dot products
between support vectors in a high-dimensional space, so there is no direct way
to ask which predictors drove a given decision. The Random Forest builds an
ensemble from hundreds of decision trees — individually interpretable structures
— and uses their average to produce stable predictions. As a by-product of this
structure, it generates variable importance scores (Figure 6), which tell us
which predictors were most useful across the ensemble. In a healthcare setting
where knowing *why* a model raises a flag is as important as whether it raises
it, this is a practical advantage the SVM simply cannot match.

**Bias–variance tradeoff.** The two models control variance in different
ways. With $C = 1$, the SVM deliberately allows some misclassifications to
keep the boundary wide and stable — a margin-geometry approach to variance
control. The Random Forest instead averages `r rf_model$ntree` de-correlated
trees, driving ensemble variance towards $\rho\sigma^2$ while keeping bias
close to that of a full unpruned tree. On mixed-predictor tabular data like
this dataset, ensemble averaging tends to be the more effective strategy, which
is reflected in the RF's stronger **`r rf_best_metric`** here. The **`r w_spec`** leads on Specificity
(`r top_spec` vs `r bot_spec` for the `r l_spec`), clearing healthy patients
more reliably. On AUC-ROC the **`r w_auc`** edges ahead (`r top_auc` vs
`r bot_auc`). In practice, the choice between the two comes down to clinical
priorities: a mass-screening programme, where false negatives are most costly,
would favour higher Sensitivity; a resource-constrained follow-up pathway,
where unnecessary referrals need to be limited, would weigh Specificity more
heavily.

**Computational scalability.** With $\mathcal{O}(n^2)$–$\mathcal{O}(n^3)$
complexity, the SVM's training time grows rapidly with dataset size. Here both
models used the full `r format(nrow(train_set), big.mark=",")` training records,
keeping the comparison on equal footing. At the scale of the full CDC BRFSS
survey (>300,000 records), however, the SVM would require subsampling or an
approximate kernel method such as Nyström approximation, while the Random
Forest could be trained in parallel on the complete dataset. This scalability
gap is independent of any single metric and is a real practical consideration
for population-level deployment.

### 5.2 Feature Importance and Key Risk Factors

Figure 6 ranks all predictors by their average contribution to node purity
across the 500 trees. The top three — **`r top3_vars[1]`**,
**`r top3_vars[2]`**, and **`r top3_vars[3]`** — each have a plausible
biological mechanism behind them:

1. **`r top3_vars[1]`**: `r var_explanation[[top3_vars[1]]]`

2. **`r top3_vars[2]`**: `r var_explanation[[top3_vars[2]]]`

3. **`r top3_vars[3]`**: `r var_explanation[[top3_vars[3]]]`

The remaining predictors show coherent signals, as the EDA confirms. One
methodological caveat worth noting: Gini importance is known to favour
continuous variables and those with many distinct values (such as `AgeCategory`
and `BMI`). The rankings in Figure 6 should therefore be read alongside the
bivariate patterns in Figures 1–3 rather than taken at face value in isolation.

### 5.3 Limitations

**Survey self-report bias.** Every variable in the dataset, including the
disease outcome itself, comes from a telephone interview. Respondents may not
recall a diagnosis accurately, and conditions that were never formally diagnosed
would not appear at all. Variables like `MentalHealth` and `PhysicalHealth`
are 30-day self-assessments of subjective wellbeing, which are not the same as
clinical measurements.

**Reverse causation.** High scores on `MentalHealth` and `PhysicalHealth`
are associated with heart disease in this data, but the direction of that
relationship is not clear. It is equally plausible that poor health days are a
symptom of living with heart disease as it is that they are a risk factor. A
cross-sectional survey cannot answer this question; a prospective cohort design
would be needed.

**Class imbalance.** The `r imbalance_ratio`:1 class ratio makes it harder
for models to learn from the minority class. Both models here were run with
default settings and no resampling. Techniques such as SMOTE, class-weighted
training, or threshold adjustment could improve minority-class detection and
would be worth exploring in follow-up work.

**SVM scalability.** On this dataset both models trained on the same
observations. At larger scale — such as the full BRFSS survey — the SVM's
quadratic complexity would require either subsampling or an approximate kernel
method, introducing a potential data-volume asymmetry that could influence
results. This is worth keeping in mind for any attempt to replicate this
analysis on the full dataset.

## 6. Conclusions

Two classification methods were compared — a **Support Vector Machine with a
Radial Kernel** (James et al., 2021, Ch. 9) and a **Random Forest** (James et
al., 2021, Ch. 8) — applied to a BRFSS health survey to identify what drives
heart disease risk.

Neither model dominates on every metric. The **`r w_sens`** catches more true
disease cases (Sensitivity = `r top_sens`), which matters most in screening
contexts. The **`r w_spec`** clears healthy patients more reliably (Specificity
= `r top_spec`). On overall discrimination, the **`r w_auc`** leads with an
AUC of `r top_auc`. Both models clear the naive baseline by a meaningful margin,
which confirms they are picking up genuine patterns rather than just predicting
the majority class. The Random Forest has the additional benefit of producing
variable importance scores — something the SVM cannot offer — which allows
clinicians to understand which factors are driving the model's output.

Age, BMI, and physical health burden emerged as the strongest predictors,
consistent with cardiovascular epidemiology (Centers for Disease Control and
Prevention, 2022). Of these, BMI is the most directly actionable. Smoking,
ranked sixth in Gini importance but well supported by the EDA and by decades
of clinical evidence, is another modifiable target. Both warrant attention in
preventive programmes aimed at middle-aged adults before age-related risk
accumulates further.

## References

Breiman, L. (2001) 'Random forests', *Machine Learning*, 45(1), pp. 5–32. doi:10.1023/A:1010933404324.

Centers for Disease Control and Prevention (2022) *Heart disease facts*. Available at: https://www.cdc.gov/heart-disease/data-research/facts-stats/index.html (Accessed: `r format(Sys.Date(), '%d %B %Y')`).

James, G., Witten, D., Hastie, T. and Tibshirani, R. (2021) *An introduction to statistical learning: with applications in R*. 2nd edn. New York: Springer.

Vapnik, V. (1995) *The nature of statistical learning theory*. New York: Springer-Verlag.

---

# Part B — Non-Technical Summary

## What Puts Your Heart at Risk?

Heart disease claims roughly one in every four lives in the United States.
Despite its scale, many people are unaware of what specifically raises their
personal risk. Researchers analysed health survey responses from US adults
collected by the Centres for Disease Control and Prevention — participants
reported their age, lifestyle, medical history, and wellbeing, along with
whether they had been diagnosed with heart disease.

**How Was It Studied?**

Two computer-based tools were used. The first, a **support vector machine**,
works by drawing the widest possible dividing line between healthy people and
those at risk on a complex health-data map. The line curves to separate the
groups as cleanly as possible, though the tool cannot easily explain why it
drew the line where it did.

The second, a **random forest**, builds 500 decision trees each trained on a
slightly different slice of the data. Every tree votes on each new patient, and
the majority wins. Because the trees each see different data and consider
different questions, their combined answer is more reliable than any single
tree. This tool also ranks which factors it found most informative.

**What Did We Find?**

Both tools identified genuine patterns well beyond random chance. Each has a
practical edge: one catches more people who truly have heart disease (fewer
missed diagnoses), while the other more accurately clears those who do not.
In healthcare, missing a real case is the graver error, so the tool with fewer
missed diagnoses is preferred for screening.

Five factors stood out as the strongest predictors:

1. **Age** is the most powerful factor — risk rises steeply and consistently
   as the body ages.

2. **BMI (body weight)** is a strong predictor. Higher weight raises blood
   pressure, worsens cholesterol, and drives inflammation that damages the heart.

3. **Physical health burden** matters. People reporting more days of poor
   physical health were significantly more likely to have heart disease,
   reflecting how chronic illness burdens the cardiovascular system over time.

4. **Sleep duration** plays a surprising role. Both too little and too much
   sleep are linked to inflammatory processes that elevate heart disease risk —
   an often overlooked but important modifiable factor.

5. **Mental health** is also significant. People who reported more days of
   poor mental wellbeing had higher rates of heart disease, reflecting the
   biological links between chronic stress, inflammation, and cardiovascular
   damage.

**Why Does This Matter?**

BMI is directly modifiable, and smoking — ranked sixth but with established
causal pathways — is also preventable. Helping people reach a healthier weight
and quit smoking, especially in their 40s and 50s, could meaningfully cut new
cases. The role of sleep and mental health as significant predictors shows that
prevention must go beyond traditional cardiac risk factors — poor sleep and
chronic stress deserve attention in routine health checks.

*Ageing cannot be stopped, but understanding how it interacts with lifestyle
gives us a real chance to intervene sooner — and save more lives.*

